LLM Output Compliance with Handcrafted Linguistic Features --- An Experiment

### 2024-06-14 article/main.tex
incorporate review

Taking Noam Chomsky's ideas~\cite{chomsky2002syntactic} that syntax and meaning
are different things, we intuit that we can define writing style only in terms
of both syntax and meaning.
We shy away from discussing meaning, but syntax and orderly constructs like it
seem to be within our grasp.

Traditionally, linguistic style was thought to be the result of morphological
and syntactic arrangements~\cite{lugea2023stylistics}.
More recent schools of thought on linguistic style find semantics and pragmatics
to be inextricably linked to author style~\cite{verma2019lexical}, but argue
that a great deal can still be inferred from lexicon, syntax and a `surface'
level which contains basic HLFs such as the number of words in a sentence.

Some have observed that many characteristics of written text can be computed as
HLFs~\cite{hovy1987generating,lugea2023stylistics}.
Combining multiple HLFs that are revelatory of linguistic style could provide us
with an easily interpretable and verifiable stylistic profile.

@@

Our experiment analyses LLM output when asked to generate new text based on
given input text and instructions derived from HLFs.
We compute statistics of HLFs on a corpus to derive the text generation
instructions.

To run the experiment we must select the LLMs to evaluate and finding datasets
from which we can derive corpuses.
Then there's the matter of selecting the HLFs for our experiment.

We think it's best to describe the experimental process before going over our
selection for the experiment's parameters.

@@

In the first scenario the target LLM is instructed to generate text retaining
the meaning of  the input text without any further instruction related to style
or the linguistic features of the generated text.
This step is repeated 10 times, computing each HLF in our selection for the
generated text.
The computed values make up the first scenario baselines for each HLF.

The second scenario consists in generating text from the same input text, but
with the aid of some examples from the corpus.
In this scenario, we believe that the choice of input text affects the outcome.
Therefore we propose two variants for this scenario: one with input text
extraneous to the corpus and one with input text from the corpus.

We add the examples to the initial LLM prompt and follow the same process like
in the first scenario.
We do this for each variant of the second scenario to obtain two more baselines
for each HLF for the corpus.

The number of examples is chosen so that we do not overrun the context window of
the LLM.
Some context windows are smaller than others, settling the number of examples at
three.

@@

Doing so allows writing an HLF prompt derived from each of our baseline prompts.
The HLF prompts instruct the LLM to generate text whose HLFs comply with the
previously computed statistics.
The individual (per-HLF) instructions are all are part of the same HLF prompt.
The prompts are constructed using Jinja templates which are located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/src/llm_tst_consistency/templates}{templates}}
directory of the source code used to run the experiment.
The instruction prompts corresponding to our two scenarios are
`\texttt{prompt\_1}' and `\texttt{prompt\_2}', respectively.
The rest of the templates in the directory are used to construct the individual
HLF instructions.

In a similar manner to how we generated the baselines, we run 10 trials for each
scenario and variant.
We compare each baseline with the results obtained by using the corresponding
HLF prompt.

@@

\subsection{User Prompt and Text}\label{input-text}

This experiment presupposes instructing various large language models to reword
an input text so that it conforms to some HLFs.

There is an input text corresponding to each corpus that is external to that
corpus.
These texts are located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/data}{data}}
folder of the experiment's source code.

For the variant that uses an input text from the corpus, we draw the text
randomly from the corpus once, at the start of the experiment.

The input text that is external to the CNN article corpus is an extract from the
classic \textit{The Life and Opinions of Tristram Shandy, Gentleman}, by
Laurence Sterne~\cite{sterne2003life}.
The text style is very dissimilar compared to news articles.
Its linguistic features are also very unlike the average on the CNN article
corpus.
A text that is similarly divorced from the style and linguistic features of most
Yelp reviews is Oscar Wilde's \textit{Sonnet toLiberty}~\cite{wilde1909poems}.

@@

Result tables highlight the difference between plots obtained using a
baseline prompt versus the ones obtained using HLF instruction prompts.
Firstly, we measure how different baselines are from HLF-enhanced plots.
Secondly, we compare the closeness of the baseline and HLF-enhanced plots to the
target average.
Table values where the text is \textbf{bold} mark significant differences
between baseline and HLF instruction results.
\underline{Underlined} values mark situations where the result of using HLF
instructions is closer to the target than the baseline.
\textit{Italic} values mark situations where the baseline is closer to the
target.

@@

In the second scenario of the experiment we try to quantify the impact of using
HLF instructions on the output of the LLM.
There are two variants of this scenario: input text from within or from outside
the corpus.

Table~\ref{table-prompt-2-cnn-dailymail} contains results for the first variant.
The outlook is similar to the first experiment scenario.

>>>>

Taking Noam Chomsky's ideas~\cite{chomsky2002syntactic} that syntax and meaning
are different things, we intuit that we can define writing style only in terms
of both syntax and meaning.
We shy away from discussing meaning, but syntax and orderly constructs like it
seem to be within our grasp.
Traditionally, linguistic style was thought to be the result of morphological
and syntactic arrangements~\cite{lugea2023stylistics}.
More recent schools of thought on linguistic style find semantics and pragmatics
to be inextricably linked to author style~\cite{verma2019lexical}, but argue
that a great deal can still be inferred from lexicon, syntax and a `surface'
level which contains basic HLFs such as the number of words in a sentence.
Some have observed that many characteristics of written text can be computed as
HLFs~\cite{hovy1987generating,lugea2023stylistics}.
Combining multiple HLFs that are revelatory of linguistic style could provide us
with an easily interpretable and verifiable stylistic profile.

@@

Our experiment analyses LLM output when asked to generate new text based on
given input text and instructions derived from HLFs.
We compute statistics of HLFs on a corpus to derive the text generation
instructions.
To run the experiment we must select the LLMs to evaluate and finding datasets
from which we can derive corpuses.
Then there's the matter of selecting the HLFs for our experiment.
We think it's best to describe the experimental process before going over our
selection for the experiment's parameters.

@@

In the first scenario the target LLM is instructed to generate text retaining
the meaning of the input text from outside the corpus without any further
instruction related to style or the linguistic features of the generated text.
This step is repeated 10 times, computing each HLF in our selection for the
generated text.
The computed values represent the baseline HLF values for the first scenario.

The second scenario consists in generating text based on input text, but with
the aid of some examples from the corpus.
The examples are appended to the instruction prompt used in the first scenario.
The number of examples is chosen so that we do not overrun the context window of
the LLM.
Some context windows are smaller than others, settling the number of examples at
three.
Adding examples to the LLM's instruction prompt also makes the provenance of the
input text more important.
Therefore we experiment with two variants of the second scenario: one with input
text extraneous to the corpus and one with input text from the corpus.
The text from outside the corpus is the same text we used in the first scenario.
The text from within the corpus is chosen randomly once at the start of the
experiment.
Following the process from the first scenario results in obtaining baseline HLF
values for each variant of the second scenario.

From here on we refer to `baseline HLF values' as `baseline'.
We now have three baselines: one for the first scenario and one for each variant
of the second scenario.

@@

These statistics allow writing an HLF prompt derived from each of our baseline
prompts.
The HLF prompts instruct the LLM to generate text with HLFs that comply with the
previously computed statistics.
The individual (per-HLF) instructions are all are part of the same HLF prompt.
The prompts are constructed using Jinja templates which are located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/src/llm_tst_consistency/templates}{templates}}
directory of the source code used to run the experiment.
The instruction prompts corresponding to our two scenarios are
`\texttt{prompt\_1}' and `\texttt{prompt\_2}', respectively.
The rest of the templates in the directory are used to construct the individual
HLF instructions.

In a similar manner to how we generated the baselines, we run 10 trials for each
scenario and variant.
We compare each baseline with the results obtained by using the corresponding
HLF prompt.
From here on we refer to the results obtained by using prompts with HLF
instructions as `HLF-enhanced results', `HLF results' or even simply `results'.

@@

\subsection{External Input Texts}\label{input-text}

The experiment involves using input text from both outside the corpus and from
within it.
This section describes how we use text external to the corpus, having already
described our usage of text from within the corpus.

We use a different external input text for each corpus.
These texts are located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/data}{data}}
folder of the experiment's source code.

The text external to the CNN article corpus is an extract from the classic
\textit{The Life and Opinions of Tristram Shandy, Gentleman}, by Laurence
Sterne~\cite{sterne2003life}.
The text style is very dissimilar compared to news articles.
Its linguistic features are also very unlike the average values computed for the
CNN article corpus.
A text that is similarly divorced from the style and linguistic features of most
Yelp reviews is Oscar Wilde's \textit{Sonnet toLiberty}~\cite{wilde1909poems}.

@@

Result tables highlight the difference between the results obtained using a
baseline prompt versus the ones obtained using HLF instruction prompts.
Firstly, we measure how different baselines are from HLF-enhanced plots.
Secondly, we compare the closeness of the baseline and HLF-enhanced plots to the
target average.
Table values where the text is \textbf{bold} mark significant differences
between baseline and HLF instruction results.
\underline{Underlined} values mark situations where the result of using HLF
instructions is closer to the target than the baseline.
\textit{Italic} values mark situations where the baseline is closer to the
target.

@@

We attempt quantifying the impact of using HLF instructions on the output of the
LLM by eliminating as many factors as possible that would lead to a different
text being generated when compared to a baseline.
In the second scenario of our experiment we use examples to allow the LLM to
generate better baselines, thus hopefully narrowing the gap between baseline
and HLF-enhanced result.

There are two variants of this scenario: input text from within or from outside
the corpus.
Table~\ref{table-prompt-2-cnn-dailymail} contains results for the first variant:
input from the corpus.
We notice these results are similar to those in the first experiment scenario.
